---
title: "YukthaPenumala-Team3-Step1"
output:
  pdf_document: default
  html_document: default
date: "2023-11-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set Working Directory
```{r}
setwd("C:/Users/ypenu/OneDrive/Documents/HIDS_7003/Final Project/Step 1")
```

## Step 0 - Check processed data 

## Step 1 - Read in data
```{r}
clinData <- read.table(file = "input/20191105_BC_ClinData_233rows.csv",
                       sep=",", 
                       header = T, 
                       stringsAsFactors = F, row.names = NULL)
```

```{r}
geneExp <- read.csv(file = "input/20191105_BC_GeneExpData_withAnno_233.tsv",
                      sep="\t",
                      row.names = 1,
                      stringsAsFactors = F)
```

## Step 2 - Filter data 

# Filter clinical data (not needed)
- Columns are samples/patients
- Rows are clinical attributes
```{r}
dim(clinData)
```

# Filter gene expression data (not needed)
- Columns are samples/patients
- Rows are molecular features (gene expression)
```{r}
dim(geneExp)
```

```{r}
matchingSamples <- which(colnames(geneExp) %in% clinData$GSMid) # 233 IDs matching
subsetGeneExp <- geneExp[, matchingSamples] ##  43148 rows features, 233 patient columns
```

## Step 3 - Identify the groups to be compared (Baseline and Comparision Groups)

```{r}
baselineGrpLabels <-  which(clinData$PrimaryBladderCancerType == "Normal bladder mucosae") #10 samples
head(baselineGrpLabels)
# Labels (row numbers) that can identify the baseline group patients
# Use the labels (row numbers) to subset baseline patients in clinical data file 
clinBase <- clinData[baselineGrpLabels, ]

# Labels (row numbers) that can identify the comp group patients
compGrpLabels <- which(clinData$PrimaryBladderCancerType == "Primary_BC_Invasive") #62 samples
head(compGrpLabels)
# Use the labels (row numbers) to subset comp patients in clinical data file 
clinComp <- clinData[compGrpLabels, ]

#### Use the clinBase and clinComp objects to subset gene expression data
geneExpTumorBase <- subsetGeneExp[, clinBase$GSMid] # 43148 feature (rows), 10 samples columns
geneExpTumorComp <- subsetGeneExp[, clinComp$GSMid] # 43148 feature (rows), 62 samples columns
```

## Step 4: Sanity check

```{r}
#See if sample ids in clinical data match sample ids in gene exp data
clinBase$GSMid == colnames(geneExpTumorBase)
clinComp$GSMid == colnames(geneExpTumorComp)
```

```{r}
# a) A tsv file that contains exported patient ids from clinical data (baseline and comparison groups)
write.table(clinBase$SUBJECT_ID , file = "output/YukthaPenumala_ClinBaseIDs.tsv", sep="\t", quote = F)
write.table(clinComp$SUBJECT_ID , file = "output/YukthaPenumala_ClinCompIDs.tsv", sep="\t", quote = F)

# b) A tsv file that contains exported patient ids from gene exp data (baseline and comparison groups)
write.table(colnames(geneExpTumorBase) , 
            file = "output/YukthaPenumala_GeneExpBaseIDs.tsv", 
            sep="\t", quote = F )
write.table(colnames(geneExpTumorComp) , 
            file = "output/YukthaPenumala_GeneExpCompIDs.tsv", 
            sep="\t", quote = F)

# c) Excel file that contains comparison on patient ids from clinical and gene exp data (baseline and comparison groups)
# Refer to file w/ name: "YukthaPenumala_Checking.xlx"

# d) A tsv file that contains exported feature names
     # This file contains only probe/gene names and no other garbage
write.table(rownames(geneExp), file = "output/YukthaPenumala_FeatureIDs.tsv", sep="\t", quote = F)
```

## Step 5 - Prep data for T-test

```{r}
### Checking to make sure data is a numeric data frame
knitr::kable(head(geneExpTumorBase[1:5,1:4]))
knitr::kable(head(geneExpTumorComp[1:5,1:4]))
```

#### Function for T-test

```{r ttest}
source("fnTTest.R")

#### Call T-test function
results1 = fnTTest(baseGroup = geneExpTumorBase, 
                   compGroup = geneExpTumorComp, 
                   testName = "TTest", 
                   baseGroupName = "Normal",
                   compGroupName =  "Invasive",
                   folderName = "output")
```

# Read in the T-Test results file
```{r}
#Read in the T-Test results file
ttestResults <- read.csv(file = "output/TTest_Invasive_(Comp).vs._Normal_(Base).TTest.csv")
```

## Step 6 - Filtering to obtain top DEGs
```{r}
#check to make sure p-value column is imported as numeric 
#sort by p-value (just in case the results are not sorted by p-value)
ttestResultsSorted <- dplyr::arrange(ttestResults, Pvalue)

#Find rows based on cut-off criteria 
whichSig <- which(ttestResultsSorted$FDR <= 0.0001 & ttestResultsSorted$SignedFC <= 1.5)

#Short list sig results
ttestResultsSig <- ttestResultsSorted[whichSig, ] #735 rows 

colnames(ttestResultsSig) # check column names of gene list

write.csv(x=ttestResultsSig, file = "output/YukthaPenumala-Invasive-Normal-DEGs-before.csv", quote=F,  row.names=FALSE)

# Split gene annotations to include only gene names. 
geneBefore <- ttestResultsSig$Feature
split_names <- strsplit(geneBefore, "\\|")
geneName <- sapply(split_names, function(x) x[2])
ttestResultsSig$Feature <- geneName
head(ttestResultsSig) # Changed gene annotations to include only gene names

### Export short listed results
write.csv(x=ttestResultsSig, file="output/YukthaPenumala_Invasive_Ttest_Shortlisted.csv", quote=F,row.names=FALSE)

# Remove any NA values as they indicate candidate genes 
ttestResultsSig2 <- na.omit(ttestResultsSig) 

length(ttestResultsSig2$Feature) #639 DEGs after removing NA values 

duplicated(ttestResultsSig2$Feature) # 6 duplicates 

# Remove 6 duplicates (keeping first occurence)
ttestResultsSig3 <- ttestResultsSig2[!duplicated(ttestResultsSig2$Feature), ]

length(ttestResultsSig3$Feature) #633 DEGs after removing NA values

### Export final short listed results
write.csv(x=ttestResultsSig3, file="output/YukthaPenumala_Invasive_Ttest_FinalShortlisted.csv", quote=F,row.names=FALSE)
```